---
title: "First steps in R"
author: "Claudius Gräbner-Radkowitsch"
date: "2/12/2022"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    theme: readable
    highlight: tango
    toc: yes
    toc_depth: 2
    number_sections: yes
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = "#>", warning = F, message = F)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tufte)
library(bit64)
```

In this post we will learn about the basic *syntax* of R.
The syntax basically refers to the grammatical rules you must adhere to when communicating with your computer in the language `R`: if you you do not follow the right syntax, i.e. you 'speak' gramatically incorrect, your computer will not understand you and communicate this to you by throwing up an error message.

To learn about these important basics, the post follows the following structure:

* Commands
* Objects, functions, and assignments
* Kind of objects


# Issue commands to your computer

There are two ways we can communicate with our computer in 
R Studio: either issuing commands directly via the *console*, or by executing a script.

Lets start by using the console and use R as a simple calculator first: we first want to add the numbers `2`and `5`.
To this end, simply type `2 + 5` into the console and press `Enter`. 
Since the expression `2 + 5` is syntactically correct R code, the 
computer 'understands' what we want from it and returns the result:

```{r}
2 + 5
```

The `#>` at the beginning of the line indicates that what is written on this line is the output of an R command (but the concrete sign might be different on your computer). 

The result of `2+5` is a number (more precisely: a 'scalar'). 
In R, scalars are always represented as a *vector* of length 1.
The `[1]` here indicates that the first element on this line is the first element of the vector.
If the result of our calculation was a very long vector that
needs to span several lines, at the beginning of the next line R would show us the index of the first number displayed on this line.^[You may try this out by typing `1:100` into your console and see what happens: this returns a vector of length 100, which certainly will contain some line breaks.]

In this way we can use R as a simple calculator,
because for all simple mathematical operations we can use certain symbols as operators.
At this point it should be pointed out that the symbol `#` in R introduces a *comment*, that means everything in a line after `#` will be ignored by the computer and
you can make notes in the code that only help *you* (or other humans) to understand what you have written.

```{r}
2 + 5 # Addition
```

```{r}
2/2 # Division
```

```{r}
4*2 # Multiplication
```

```{r}
3**2 # Exponentiation
```

As an alternative to typing the commands into the console and then press `Enter`
to execute them, we can write down the commands in a script, and then execute 
this script.
While the interaction via the console is useful to test the effects of certain 
commands, scripts are useful if we want to develop more complex operations, and
save what you have written for later, or to make them accessible to other people:
we can save scripts as a file on our computer, and then use them later. 

The operations that we have conducted so far and not particularly exciting, to
be honest. Before we proceed with more complex operations, however, we need to
understand the ideas of `objects`, `functions`, and `assignments`.

# Objects, functions, and assignments

> To understand computations in R, two slogans are helpful:
>   Everything that exists is an object.
>   Everything that happens is a function call.
> `r tufte::quote_footer('John Chambers')`

The statement 'Everything that exists is an object.' means that every number,
function, letter, or whatever there is, is an object that is stored somewhere
in the physical memory of your computer.
For instance, in the computation 
`2 + 3`, the number `2` is as much an object as the number `3`
and the addition-function, which we call via the operator `+`.

The statement 'Everything that happens is a function call.' means that whenever
we tell our computer to do something via R, we are effectively calling a 
*function*.

**Functions** are algorithms that apply certain routines to an *input* and 
produce an *output*. 
The addition function we called in the calculation `2 + 3` took as input the 
two numbers `2` and `3`, applied to them the addition routine and produced the 
number `5` as output.
The output `5` is an object in R just like the inputs `2` and `3`, 
as well as the addition function.

A 'problem' is that in the present case R prints the output of the calculation 
but we have no access to it afterwards:

```{r}
2 + 3
```

It is stored, for some time, on the physical memory of our computer, but we
basically have no way to find it. To address this problem we can issue an 
*assignment*: whenever we want to keep using the output of an operation, we may
give the output a *name*. This name works effectively as a kind of pointer, 
which points to the place on the computer memory where the output is saved.
This way, we can access, and reuse it whenever we call the name. The process of
giving a name to an object is called *assignment*, and it is effectuated via the
function `assign`:

```{r}
assign("intermediate_result", 2 + 3)
```

We explain the process of calling a function in more detail below. Here we focus
on the process of assignment instead. What the function `assign` does is the 
following: it assigns the name `intermediate_result` to the result of the 
operation `2 + 3`. We can now call this result by writing its name into the 
console and press `Enter`:

```{r}
intermediate_result
```

Since making assignments happens so frequently in practice, there is a shortcut
to the use of the function `assign`, namely the operator `<-`.
Thus, the following two commands to effectively the same thing:

```{r}
assign("intermediate_result", 2 + 3)
intermediate_result <- 2 + 3
```

From now on, we will only use the `<-` operator, which also represents
quite nicely the idea of assignments as pointers to certain objects.^[In theory 
we can use `<-` also the other way around: `2 + 3 -> intermediate_result`.
At first sight this is more intuitive and respects the sequence of events:
first, the result of `2 + 3` gets created, i.e. a new object gets defined.
Then, this object gets the name `intermediate_result`.
However, the code that results from such practice is usually much more difficult
to read, so it is common practice to use `<-` rather than `->`.] 

> **Digression: why `<-`?** 
The use of the string `<-` as an assignment operator is at first sight 
unintuitive, uncomfortable, and rather unique in the world of programming languages. 
Much more common is the use of `=`. Where does this particularity of R come from? 
Besides practical reasons -- in contrast to `=`, the use of `<-` makes explicit
the unidirectionality of an assignment -- the main reason is historical: 
`R` originated from the programming language `S`. This in turn has taken over 
the `<-` from the language `APL`. And APL, in turn, was developed on a keyboard 
layout, where `<-` had its own key. Moreoever, since the operator `==` was not 
commonly used at that time, `=` was already assigned as test for equality 
(which, today, is basically always done by using `==`).
And so one has decided to use `<-` as an assignment operator and while since 
2001 you can also make assignments in R using `=`, `<-` remains strictly 
recommended for the sake of readability as well as some technicalities.

You are not allowed to give names to objects as you wish. All syntactically 
correct names in R...

* only contain letters, numbers, or the symbols `.` and `_`
* do not start with `.` or a number 

Moreover, there are some reserved words that you must not (and cannot) use as
names, e.g. `function`, `TRUE`, of `if`. You can have a look at the complete
list of forbidden words by calling `?Reserved`.

There is, however, nothing to remember since whenever you try to give an object
a name that conflicts with the rules just described, R immediately throws an
error message:

```{r run, error=TRUE}
TRUE <- 5
```

There are, however, some rules that determine what is a *good* name and that
you should adhere to whenever possible:

* Names should be short and informative; `sample_mean` is a good name, `vector_2` not so much
* You should **never use special characters**, especially *Umlaute*
* R is *case sensitive*, meaning that `mean_value` is a different name than `Mean_Value`
* Even if this is possible you should never use names that are already used for
function provided by R. For instance, an assignment such as `assign <- 2` 
is possible, but it effectively prevents you from using the function `assign` 
without further complications.


> **Note**: You can have a look at all current assignments in the `Environment`
pane in R-Studio, or list them by calling `ls()`

> **Note**: One object can have more than one name, but no name can ever point
to two object. If you re-assign a name, the old assignment will be overwritten:

```{r}
x <- 2 
y <- 2 # The objekt 2 now has two names
print(x)
print(y)
x <- 4 # The name 'x' now points to '4', not to '2'
print(x)
```

> **Note**: As you might have experienced, R does not return results after 
making an assignment:

```{r}
2 + 2 # No assignment, R returns the result in the console
```

```{r}
x <- 2 + 2 # Assignment, R does not return the results in the console
```

Lets recap what we have learned so far about issuing commands, names and 
assignments:

* We can issue commands to the computer in R by (a) typing R code into the 
console and press `Enter`, or (b) write the code into a script and then execute 
it
* Everything that *exists* in R is an *object*, everything that *happens* is a
function call
* A function is an object, that takes an input, applies a certain routine, and
returns an output
* We can assign an object a name by using `<-`. Then we can call this object by
typing its name. The process of giving a name to an object is called *assignment*,
and we can have a look at all names currently given to objects by calling `ls()` 
anzeigen lassen

Finally, I want to point your attention to the function `help()`, which 
can provide you with additional information about the object a name points to.
For instance, if you want to get more information about the function with the
name `assign`, then just type the following: 

```{r, eval=FALSE}
help(assign)
```
